---
title: "Statistical Computing HW 1"
author: "Mario Ibanez"
date: "January 16, 2016"
output: pdf_document
---

```{r message=FALSE, warning=FALSE}
library(knitr)
```

# Problem 3.2)   

## Question
If $x_0=3$ and
$$
x_n = (5x_{n-1}+7) \;\; mod \; 200
$$
find $x_1,...x_{10}$.   

## Answer
The code below finds $x_1,...x_{10}$:
```{r}
# Create a dataframe to store the values
df2 <- data.frame(x_i=c(0:10), equals=rep(0,11))

# Initialize x0 = 3
df2[1,2] = 3

# Loop for x1 to x10
for (i in 1:10){
  df2[i+1, 2] = (5*df2[i, 2]+7) %% 200
}
```

```{r results='asis'}
# Print results
kable(df2)
```

# Problem 3.5)   

## Question
Use simulation to approximate
$$
\int_{-2}^{2} e^{x+x^2} dx
$$
Compare answer with the exact answer if known.

## Answer
**1) Derivation**   
**2) Algorithm**   
**3) Simulation**   
**4) Analytical Result**   

# Problem 3.6)   

## Question
Use simulation to approximate
$$
\int_{0}^{\infty} x(1+x^2)^{-2} dx
$$
Compare answer with the exact answer if known.

## Answer
**1) Derivation**   
**2) Algorithm**   
**3) Simulation**   
**4) Analytical Result**   



