---
title: "Bootstrap-t Comparison"
author: "Mario Ibanez"
date: "March 26, 2016"
output: pdf_document
---

# Question
Let $x_1, x_2, ... x_{20}$ each be independent values from an exponential distribution with mean = $1/lambda$. (In other words, $f(x) = \lambda e^{-\lambda x}$).  Use the Bootstrap-t method to construct a 95% confidence interval for the parameter $\lambda$.

# Answer
Below is a random sample of 20 values from the exponential distribution:

```{r}
# Set seed
set.seed(1234)

# Load packages and data
library(ggplot2)
library(knitr)

# 20 random exponential values with mean = 1/10
lambda <- 10
data <- rexp(n = 20, rate = lambda); data
```

The 95% confidence interval will be calcuated two ways.

# Method 1
This method will use the delta method to estimate the variance of $\hat{\theta} = 1/\overline{x}$.  We have:

$$
S_{\hat{\theta}} = \frac{S}{\overline{x}^2 \sqrt{n}}
$$
$$
\hat{\theta} = \frac{1}{\overline{x}}
$$

Below are the values of $\hat{\theta}$ and $S_{\hat{\theta}}$

```{r}
# Values for original sample
theta_hat <- 1 / mean(data); theta_hat
s_theta_hat <- sd(data) / (mean(data)^2 * sqrt(length(data))); s_theta_hat
```

Now we can bootstrap 1000 samples from this data.  First, this is the function that will be used to generate the "t-statistic" for each bootstrapped sample:

```{r}
# Function to calculate statistics
statistic_generator <- function(x, theta_hat){
  theta_hat_star <- 1 / mean(x)
  s_theta_hat_star <- sd(x) / (mean(x)^2 * sqrt(length(x)))
  return((theta_hat_star - theta_hat)/s_theta_hat_star)
}
```

This function will calculate the values $\hat{theta}$, $S_{\hat{\theta}}$, and $T_i^*$ when given a bootstrapped sample.

Now 1000 bootstrapped samples are calculated:

```{r}
boot_strapped_data <- matrix(data = sample(x = data, size = 20*1000, replace = TRUE),
                             ncol = 20, byrow = TRUE)
```

Now the statistics are calculated:

```{r}
boot_statistics <- matrix(apply(X = boot_strapped_data, MARGIN = 1, 
                                FUN = statistic_generator, theta_hat = theta_hat), 
                          ncol = 1, byrow = TRUE)
```

Now we can calcuate the 95% confidence interval using the 2.5th and 97.5th percentiles of the "t-statistics":

```{r}
sorted_t_statistics <- sort(boot_statistics[, 1])
lower_limit <- theta_hat - sorted_t_statistics[1000*0.975] * s_theta_hat; lower_limit
upper_limit <- theta_hat - sorted_t_statistics[1000*0.025] * s_theta_hat; upper_limit
```

Now we can try to run this same method 100 times in order to approximate the true coverage probability of this method.

```{r}
results_vector <- rep(0, 100)
confidence_width <- rep(0, 100)

for(i in 1:100){
  boot_strapped_data <- matrix(data = sample(x = data, size = 20*1000, replace = TRUE),
                               ncol = 20, byrow = TRUE)

  boot_statistics <- matrix(apply(X = boot_strapped_data, MARGIN = 1, 
                                FUN = statistic_generator, theta_hat = theta_hat), 
                            ncol = 1, byrow = TRUE)
  
  sorted_t_statistics <- sort(boot_statistics[, 1])
  lower_limit <- theta_hat - sorted_t_statistics[1000*0.975] * s_theta_hat; lower_limit
  upper_limit <- theta_hat - sorted_t_statistics[1000*0.025] * s_theta_hat; upper_limit
  
  results_vector[i] <- ((lambda < upper_limit) && (lambda > lower_limit))
  confidence_width[i] <- upper_limit - lower_limit
}
```

Out of 100 simulations, $`r sum(results_vector)`$ contained the true value of lambda.  The average confidence interval with was $`r mean(confidence_width)`$

# Method 2
This method creates a confidence interval instead for the value $1/\lambda$ directly, in otherwords, $\hat{\theta} = \overline{x}$.  At the end, the confidence interval is transformed into a confidence interval for $\lambda$


```{r}
# Values for original sample
theta_hat2 <- mean(data); theta_hat2
s_theta_hat2 <- sd(data) / sqrt(length(data)); s_theta_hat2
```

Now we can bootstrap 1000 samples from this data.  First, this is the function that will be used to generate the "t-statistic" for each bootstrapped sample:

```{r}
# Function to calculate statistics
statistic_generator2 <- function(x, theta_hat2){
  theta_hat_star <- mean(x)
  s_theta_hat_star <- sd(x) / sqrt(length(x))
  return((theta_hat_star - theta_hat2)/s_theta_hat_star)
}
```

So that the comparison is more equivalent, the same bootstrapped data will be used.  Or course new "t-statistics" need to be calculated however.

```{r}
boot_statistics2 <- matrix(apply(X = boot_strapped_data, MARGIN = 1, 
                                FUN = statistic_generator2, theta_hat = theta_hat2), 
                          ncol = 1, byrow = TRUE)
```

Now we can calcuate the 95% confidence interval using the 2.5th and 97.5th percentiles of the "t-statistics":

```{r}
sorted_t_statistics2 <- sort(boot_statistics2[, 1])
lower_limit2 <- theta_hat2 - sorted_t_statistics2[1000*0.975] * s_theta_hat2; lower_limit2
upper_limit2 <- theta_hat2 - sorted_t_statistics2[1000*0.025] * s_theta_hat2; upper_limit2
```

These lower and upper limits are for the value of $1 / \lambda$.

Now we can try to run this same method 100 times in order to approximate the true coverage probability of this method.

```{r}
results_vector2 <- rep(0, 100)
confidence_width2 <- rep(0, 100)

for(i in 1:100){
  boot_strapped_data <- matrix(data = sample(x = data, size = 20*1000, replace = TRUE),
                               ncol = 20, byrow = TRUE)

  boot_statistics <- matrix(apply(X = boot_strapped_data, MARGIN = 1, 
                                FUN = statistic_generator2, theta_hat = theta_hat2), 
                            ncol = 1, byrow = TRUE)
  
  sorted_t_statistics <- sort(boot_statistics[, 1])
  upper_limit <- 1 / (theta_hat2 - sorted_t_statistics[1000*0.975] * s_theta_hat2)
  lower_limit <- 1 / (theta_hat2 - sorted_t_statistics[1000*0.025] * s_theta_hat2)
  
  results_vector2[i] <- ((lambda < upper_limit) && (lambda > lower_limit))
  confidence_width2[i] <- upper_limit - lower_limit
}
```

Out of 100 simulations, $`r sum(results_vector2)`$ contained the true value of lambda.  The average confidence interval with was $`r mean(confidence_width2)`$  

This method also had good performance but the confidence intervals were much wider on average.