---
title: "Statistical Computing HW 2"
author: "Mario Ibanez"
date: "January 17, 2016"
output: pdf_document
---

```{r}
set.seed(1234)
```

# Problem 3.10)

## Question
Use simulation to approximate $cov(U,e^U)$ where $U$ is uniform on $(0,1)$.  Compare your approximation with the exact answer.

## Answer
### 1) Derivation
Using the definition of covariance, we have:
$$
cov(U,e^U) = E(Ue^U)-E(U)E(e^U)
$$
No other derivation is needed.

### 2) Algorithm
1) Generate $u$ from $U(0,1)$
2) Evaluate $Ue^U$, $U$, and $e^U$ as three separate values, store values
3) Repeat (1) and (2) 10,000 times
4) Find the average of each of the the 10,000 values, then find $E(Ue^U)-E(U)E(e^U)$

### 3) Simulation

```{r}
# Initialize dataframe, 3 columns and 10,000 rows
df <- data.frame(UeU=rep(0,10000),
                 U=rep(0,10000),
                 eU=rep(0,10000))
# For loop to calcuate Ue^U, U, and e^U, and store values
for(i in 1:10000){
  U <- runif(1)
  df[i,1] <- U*exp(U)
  df[i,2] <- U
  df[i,3] <- exp(U)
}

# Find column means, then perform final calculation
covariance <- mean(df$UeU) - mean(df$U)*mean(df$eU)
covariance
```

The estimated covariance is $`r covariance`$.

### 4) Analytical Result
The exact value of the covariance is:
\begin{align*}
cov(U,e^U) 
&= E(Ue^U)-E(U)E(e^U) \\
&= \int_0^1 Ue^U \; du - \left( \int_0^1 U \; du \right) \left( \int_0^1 e^U \; du \right) \\
&= \left[ Ue^U - e^U \right]_0^1 - \left[\frac{1}{2}U^2 \right]_0^1 * \left[ e^U \right]_0^1 \\
&= (e-e)-(0-1)-(1/2 - 0)(e-1) \\
&= (3/2) - e/2 \\
&\approx 0.1408591
\end{align*}





