---
title: "Statistical Computing HW 3"
author: "Mario Ibanez"
date: "January 23, 2016"
output: pdf_document
---

# Problem 5.1)
## Question
Give a method for generating a random variable having density function
$$
f(x) = e^x / (e-1), \;\; 0 \le x \le 1
$$

## Answer

### Derivation
\begin{align*}
F(x)
&= \int_0^x \frac{1}{e-1} e^t dt \\[1ex]
&= \frac{1}{e-1} \left[ e^t  \right]^x_0 \\[1ex]
&= \frac{1}{e-1} (e^x - 1) \\[1ex]
&= \frac{e^x - 1}{e -1}
\end{align*}
Then set $U = F(x)$ where $U$ is a standard normal uniform random variable, and solve for $x$:
\begin{align*}
u &= \frac{e^x - 1}{e -1} \\[1ex]
(e-1)u + 1 &= e^x \\[1ex]
ln((e-1)u+1) &= x \\[1ex]
\end{align*}
and so we have:
$$
X = ln((e-1)U+1) \\[1ex]
$$

### Algorithm
1) Generate N values from $U(0,1)$  
2) Evaluate $x = ln((e-1)u+1)$ for each of the N u values

### Program
The R function *runif()* creates a random vector which can then be used to do element-wise operations with.  Loops in R should be avoided when possible:

```{r}
# 10,000 randomly generated values
N <- 10^5
random_x <- log((exp(1) - 1) * runif(N) + 1)
```

### Visualization
Below is a plot to confirm our results visually:

```{r}
# Histogram of random numbers along with curve of f(x) function
hist(random_x, freq = FALSE, breaks = seq.int(0,50)/50,
     xlab = "Randomly generated values from X",
     main = "Histogram of values from X" )
curve(exp(x)/(exp(1)-1), 0, 1, add = TRUE, col = "red")
```

$\pagebreak$


# Problem 5.3)
## Question
Use the inverse transform method to generate a random variable having distribution function
$$
F(x) = \frac{x^2 + x}{2}, \;\; 0 \le x \le 1
$$

## Answer

### Derivation
We have:
\begin{align*}
u &= \frac{x^2 + x}{2} \\[1ex]
2u &= x^2 + x \\[1ex]
0 &= x^2 + x - 2u \\[1ex]
x &= \frac{-1 \pm \sqrt{1 - 4(1)(-2u)}}{2} \\[1ex]
x &= \frac{-1 \pm \sqrt{1 + 8u)}}{2} 
\end{align*}
No we take only the addition case because we know that $x$ cannot be negative:
$$
x = \frac{\sqrt{1 + 8u)} - 1}{2} 
$$
and so:
$$
X = \frac{\sqrt{1 + 8U)} - 1}{2} 
$$

### Algorithm
1) Generate N values from $U(0,1)$  
2) Evaluate $x = \frac{\sqrt{1 + 8u)} - 1}{2}$ for each of the N u values

### Program
The R function *runif()* creates a random vector which can then be used to do element-wise operations with.  Loops in R should be avoided when possible:

```{r}
# 10,000 randomly generated values
N <- 10^5
random_x <- (1/2)*(sqrt(1 + 8 * runif(N)) - 1)
```

### Visualization
Below is a plot to confirm our results visually, with $f(x) = x + 1/2$ (note that the distribution is a trapezoid):

```{r}
# Histogram of random numbers along with curve of f(x) function
hist(random_x, freq = FALSE, breaks = seq.int(0,50)/50,
     xlab = "Randomly generated values from X",
     main = "Histogram of values from X" )
abline(0.5, 1, col = "red")
```

