---
title: "Statistical Computing HW 1"
author: "Mario Ibanez"
date: "January 16, 2016"
output: pdf_document
---

```{r message=FALSE, warning=FALSE, echo=FALSE}
library(knitr)
```

```{r}
## Set seed
set.seed(1234)
```

# Problem 3.2)

## Question
If $x_0=3$ and
$$
x_n = (5x_{n-1}+7) \;\; mod \; 200
$$
find $x_1,...x_{10}$.   

## Answer
The code below finds $x_1,...x_{10}$:
```{r}
# Create a dataframe to store the values
df2 <- data.frame(x_i=c(0:10), equals=rep(0,11))

# Initialize x0 = 3
df2[1,2] = 3

# Loop for x1 to x10
for (i in 1:10){
  df2[i+1, 2] = (5*df2[i, 2]+7) %% 200
}
```

```{r results='asis'}
# Print results
kable(df2)
```

# Problem 3.5)   

## Question
Use simulation to approximate
$$
\int_{-2}^{2} e^{x+x^2} dx
$$
Compare answer with the exact answer if known.

## Answer
### 1) Derivation   
A change of variables is needed.  Let:
$$
y = \frac{x+2}{4}
$$
This is a one to one (increasing) function that maps the interval $(-2,2)$ to $(0,1)$.  Then:
\begin{align*}
x &= 4y-2 \\
dx &= 4 \; dy
\end{align*}
Substituting these values in the integral we get:
$$
\int_{-2}^{2} e^{x+x^2} dx = \int_0^1 e^{4y-2+(4y-2)^2}(4) dy
$$
Let:
$$
h(y) = e^{4y-2+(4y-2)^2}(4) dy
$$
Then:
\begin{align*}
\int_0^1 e^{4y-2+(4y-2)^2}(4) dy 
&= \int_0^1 h(y) dy \\
&= \int_0^1 h(y)(1) dy \\
&= E(h(y)), \;\; Y \sim U(0,1)
\end{align*}
The final integral is then the expected value of $h(y)$, where $Y$ is a standard uniform random variable.   

### 2) Algorithm

1) Generate $y$ from $U(0,1)$
2) Evaluate $h(y)$, store value
3) Repeat (1) and (2) 100000 times
4) Find the average of the 100000 values

$\pagebreak$

### 3) Simulation

```{r}
# initialize a vector of length 10000
vector <- rep(0, 100000)

# loop to generate y values and evalute h(y)
for(i in 1:100000){
  # generate random number from U(0,1)
  y <- runif(1)
  # evalute and store h(y)
  vector[i] <- exp(4*y-2+(4*y-2)^2)*4
}

# compute the average
average <- mean(vector)
average
```

The average value of the 100000 $h(y)$ values is $`r average`$.

### 4) Analytical Result

To my knowledge, there is no simple way to integrate this function.  WolframAlpha provides an estimate of 93.1628 however.

# Problem 3.6)   

## Question
Use simulation to approximate
$$
\int_{0}^{\infty} x(1+x^2)^{-2} dx
$$
Compare answer with the exact answer if known.

## Answer
### 1) Derivation
First, it worthwhile to check that this integral converges.  This will be done in Part 4 of the answer, and it will be shown that it does in fact converge.  To use simulation, a change of variables is needed. Let:
$$
y = \frac{x}{x+1} = 1 - \frac{1}{x+1}
$$
This is a one to one (increasing) function that maps the interval $(0,\infty)$ to $(0,1)$.  Then:
\begin{align*}
x &= \frac{y}{1-y} \\
dx &= \frac{1}{(1-y)^2} dy
\end{align*}
Then making the change of variables, we have:
\begin{align*}
\int_{0}^{\infty} x(1+x^2)^{-2} dx
&= \int_{0}^{1} \frac{y}{1-y}\left(1+\left(\frac{y}{1-y}\right)^2\right)^{-2} \frac{1}{(1-y)^2} dy \\
&= \int_{0}^{1} \frac{y}{(1-y)^3 (1+(\frac{y}{1-y})^2)^2} dy
\end{align*}
Then let:
$$
h(y) = \frac{y}{(1-y)^3 (1+(\frac{y}{1-y})^2)^2}
$$
Then:
\begin{align*}
\int_{0}^{1} \frac{y}{(1-y)^3 (1+(\frac{y}{1-y})^2)^2}
&= \int_{0}^{1} h(y) dy \\
&= \int_0^1 h(y)(1) dy \\
&= E(h(y)), \;\; Y \sim U(0,1)
\end{align*}
The final integral is then the expected value of $h(y)$, where $Y$ is a standard uniform random variable. 

### 2) Algorithm

1) Generate $y$ from $U(0,1)$
2) Evaluate $h(y)$, store value
3) Repeat (1) and (2) 100000 times
4) Find the average of the 100000 values

### 3) Simulation

```{r}
# initialize a vector of length 10000
vector <- rep(0, 100000)

# loop to generate y values and evalute h(y)
for(i in 1:100000){
  # generate random number from U(0,1)
  y <- runif(1)
  # evalute and store h(y)
  vector[i] <- y / ((1-y)^3 * (1+(y/(1-y))^2)^2)
}

# compute the average
average <- mean(vector)
average
```

The average value of the 100000 $h(y)$ values is $`r average`$.

### 4) Analytical Result

The exact value of the integral can be found (using a change of variable: $y=x^2$, $dy=2x\;dx$):

\begin{align*}
\int_{0}^{\infty} x(1+x^2)^{-2} dx
&= \int_{0}^{\infty} \frac{x}{(1+x^2)^{2}} dx \\
&= \frac{1}{2} \int_{0}^{\infty} \frac{2x}{(1+x^2)^{2}} dx \\
&= \frac{1}{2} \int_0^\infty \frac{1}{(1+y)^{2}} dy \\
&= -\frac{1}{2} \left[ \frac{1}{1+y} \right]^{\infty}_0 \\
&= -\frac{1}{2} (0-1) \\
&= \frac{1}{2}
\end{align*}

The exact value of the integral is $\frac{1}{2}$.



